---
title: "Assignment 5"
author: "Ethan"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  message: false
  warning: false
---

# Set Up

## Packages 

```{r}
#| label: packages

library(tidyverse)
library(here)
library(janitor)
library(snakecase)

#Wrap Axis Label
library(scales)

#reading .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# visualizing
library(patchwork)
```


## Data 
```{r}
#| label: data

# taxnomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality

water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)
```

## Necessary Code from Class

```{r}
#| label: ncos-sites

# creating new object from sites
ncos_sites <- sites %>% 
  # filering to only include sites at NCOS, sampled quarterly
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")



# creating new object from taxon_list
taxon_list_clean <- taxon_list %>% 
  # converted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))



water_quality_clean <- water_quality %>% 
  clean_names() %>% 
  semi_join(ncos_sites, by = "site") %>% 
  unite("date_site", date, site, remove = TRUE) %>%
  group_by(date_site) %>% 
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) %>% 
  ungroup() %>% 
  filter(date_site != "2022-08-12NVBR")


aq_ins_clean <- aq_ins %>% 
  clean_names() %>% 
  semi_join(ncos_sites, by = "site") %>% 
  rename(date = date_on_vial) %>% 
  select(date, sample_type,site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae,
         annelida_oligochaete,
         annelida_polychaete,
         diptera_chironomid) %>% 
  filter(sample_type %in% c("FB250", "FB 250")) %>% 
  pivot_longer(
    # columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance"
  ) %>% 
  group_by(date, site, taxon) %>% 
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) %>% 
  ungroup() %>% 
  unite("date_site", date, site, remove = FALSE) %>% 
  left_join(water_quality_clean, by = "date_site") %>% 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) %>% 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "East Channel",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NDC" ~ "Devereux Creek")) %>% 
  mutate(site_full = fct_reorder(site_full,
                                 med_sal,
                                 .fun = "median",
                                 .na_rm = TRUE),
         site_full = fct_rev(site_full))


copepods <- aq_ins_clean %>% 
  filter(taxon == "copepod") %>% 
  mutate(log_ave_lit = log(ave_lit + 1))

```


# 1. Visualizing differences across sites in major taxon

## a. Planning the Figure

- x-axis: Salinity (ppt)
- y-axis: taxon
- geometry to represent average abundance/L: geom_jitter()
- geometry to represent medians: stat_summary()

## b. Wrangling the data

```{r}
#| label: ostracod-wrangling

ostracods <- aq_ins_clean %>% 
  filter(taxon == "ostracod") %>% 
  mutate(log_ave_lit = log(ave_lit + 1))

slice_sample(ostracods, n = 10)

```


## c. Coding the figure

```{r}
#| label: ostracod-visualization

ostracod_plot <- ggplot(data = ostracods,
                       mapping = aes(x = site_full,
                                     y = log_ave_lit))+
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  stat_summary(geom = "point",
              fun = median,
              color = "green",
              size = 2)+
  scale_x_discrete(labels = label_wrap(width = 10)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 8)) +
                     labs(x = "site",
                          y = "log + 1 transformed average abundance/L",
                          title = "Ostracods") +
                     theme_bw()
ostracod_plot
```


## d. Creating a multipanel figure


```{r}
#| label: plotting-objects-from-class

# creating an object of a salinity plot
salinity_plot <- ggplot(data = aq_ins_clean,
                        mapping = aes(x = site_full,
                                      y = med_sal))+
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21)+
  stat_summary(geom = "point",
               fun = median,
               color = 'red')+
  scale_x_discrete(labels = label_wrap(width = 10))+
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 200)) +
  labs(x = "site",
       y = " Median Salinity (ppt)",
       title = "Site Salinity")



# creating a plotting object of copepod abundance
copepod_plot <- ggplot(data = copepods,
                       mapping = aes(x = site_full,
                                     y = log_ave_lit))+
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  stat_summary(geom = "point",
              fun = median,
              color = "blue",
              size = 2)+
  scale_x_discrete(labels = label_wrap(width = 10)) +
scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 10)) +
                     labs(x = "site",
                          y = "log + 1 transformed average abundance/L",
                          title = "Copepod") +
                     theme_bw()

```

```{r}
#| label: creating-multi-pane
#| fig-width: 12
#| fig-height: 8

salinity_plot /
  (copepod_plot | ostracod_plot)
```


## e. Interpreting the figure

The biggest differences in copepod and ostracod abundance across sites can be seen at the East Channel site, where copepods are highest in abundance, and osrtacods are extremely low in abundance. The two species are similar in abundance at the Main Channel, South of Phelps Bridge, North of Phelps Bridge, and Devereux Creek sites.

There doesn't appear to be a correlation between site salinity and copepod or ostracod abundance. Median salinity is order by descending ppt from left to right on the x-axis, and the invertebrate abundance plots with the same x-axis don't follow a descending or ascending pattern, appearing relatively random.


# 2. Visualizing total abundance as a function of salinity



